---
title: "Collecting Polynomial Terms"
author: "Doug Hemken"
date: "November 5, 2018"
output: html_fragment
---

```{r, echo=FALSE}
source("stdParm functions.r")
```
When using Kronecker products to form polynomial transformations, we want 
to collect like terms.  That is, if I have a variable $x$ with mean
$\mu$ in an equation
$$y = \beta_0 + \beta_1 x + \beta_2 x^2$$
that I would like to recenter, with $x_\delta = x - \mu$ so that
$$y = \beta_{\delta 0} + \beta_{\delta 1}x_\delta + \beta_{\delta 2} x_\delta^2$$
The basic recentering matrix can be formed in the same manner as
for interaction terms.

```{r}
mu <- 3
names(mu) <- "x"
A <- mean.to.matrix(mu) # recentering x - 3
C <- kron(A,A)
C
```

However, notice that we have two $x$ terms in the rows and columns.  This
implies that our original equation is of the form
$$y = b_0 + b_1 x + b_2 x + b_3 x^2$$
We can rewrite our original equation in this form, letting $b_2=0$ - the
$x$ terms were already collected in our original equation.

Now notice that this coefficient choice zeros out one column of our
recentering matrix.  We could write this more succinctly by dropping
the zeroed column and leaving our coefficient vector in it's original form.

```{r}
cnames <- colnames(C)
C <- C %*% diag(1,4,4)[,-3]
colnames(C) <- cnames[-3]
C
```

We still have two rows for the $x_\delta$ terms. These do ***not*** zero
out, so we want to collect them in a single term.

```{r}
rnames <- rownames(C)
C <- cbind(c(1,0,0),c(0,1,0),c(0,1,0),c(0,0,1)) %*% C
rownames(C) <- rnames[-3]
C
```