---
title: "Collecting Polynomial Terms"
author: "Doug Hemken"
date: "`r Sys.Date()`"
output:
  html_document:
    highlight: null
  pdf_document:
    highlight: null
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(comment=NA)
library(stdParm)
```

## Recentering
When using Kronecker products to form polynomial transformations, we want 
to collect like terms.  That is, if I have a variable $x$ with mean
$\mu$ in an equation
$$y = \beta_0 + \beta_1 x + \beta_2 x^2$$
that I would like to recenter, with $x_\delta = x - \mu$ so that
$$y = \beta_{0}^\delta  + \beta_{1}^\delta x_\delta + \beta_{2}^\delta  x_\delta^2$$
The basic recentering matrix can be formed in the same manner as
for interaction terms.

```{r kronecker}
b <- c(1,1,1)
b.terms <- names(b) <- c("(Intercept)", "x", "x:x")

mu <- 3
names(mu) <- "x"

A <- mean.to.matrix(mu) # recentering x - 3
C <- kron(A,A)
C
```

However, notice that we now have two $x$ terms in the rows and columns.  This
implies that our original equation is of the form
$$y = b_0 + b_1 x + b_2 x + b_3 x^2$$
We can rewrite our original equation in this form, letting $b_2=0$, recognizing the
$x$ terms were already collected in our original equation.

Now notice that this coefficient choice zeros out one column of our
recentering matrix.  We could write this more succinctly by dropping
the zeroed column and leaving our coefficient vector in it's original form.

```{r columns}
cnames <- colnames(C)
keepcols <- match(b.terms, cnames)
C <- C[, keepcols]
colnames(C) <- cnames[keepcols]
C
```

We still have two rows for the $x_\delta$ terms. These do ***not*** zero
out, so we want to collect them in a single term.

```{r collect}
cnames <- colnames(C) # reduced columns
rnames <- rownames(C)
rowcombine <- matrix(NA, nrow=length(cnames), ncol=length(rnames))
for (i in 1:length(cnames)) {
  rowcombine[i,] <- rnames %in% cnames[i]
}
C <- rowcombine %*% C
rownames(C) <- cnames
C
```

All of this is packaged in the function `polyterm()`
```{r polyterm}
polyterm(mu, b.terms)
```

A cubic example
```{r cubic}
b <- c(1,1,1,1)
b.terms <- names(b) <- c("(Intercept)", "x", "x:x", "x:x:x")

mu <- 3
names(mu) <- "x"

polyterm(mu, b.terms)
```

## Rescaling
If we are rescaling, no terms need be collected.  Rescaling constants are
multiplied, the same as if these polynomial terms were interaction terms.

```{r polyscale}
b <- c(1,1,1,1)
b.terms <- names(b) <- c("(Intercept)", "x", "x:x", "x:x:x")

sigma <- 3
names(sigma) <- "x"

polyterm(sigma, b.terms, type="scale")